---
title: "ReproducibleResearch_Project2_WeatherEvents"
author: "Dongying"
date: "8/11/2020"
output: html_document
---
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.   
In the following paragraphs, I'll take a glance of the whole dataset and focus on damages to population health and economy due to the weather events.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading the dataset and quick summary
First of all, I download the data using the URL and read it into R as a data frame.
```{r}
filename <- "repdata_data_StormData.csv.bz2"
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists(filename)){
        download.file(fileURL, filename, method = "curl")
}
stormData <- read.csv(filename, sep = ",", header = TRUE)
```
To get a first impression of the data, we can look at the column names to see what kind of data do we have by calling the "names()" function:
```{r}
names(stormData)
```
Also, we can see the summary of it:
```{r}
summary(stormData)
```

## Questions
### Q1: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
To answer this question, we first need to extract the relevant data. By looking through the column names, we can see only "FATALITIES" and "INJURIES" are related to population health. So we can sum for those numbers by event types:
```{r}
fatal_byType <- aggregate(FATALITIES ~ EVTYPE, stormData, sum)
injury_byType <- aggregate(INJURIES ~ EVTYPE, stormData, sum )
```
Sort the data in a descending order:
```{r}
fatal_byType_Sorted <- fatal_byType[order(-fatal_byType$FATALITIES),]
injury_byType_Sorted <- injury_byType[order(-injury_byType$INJURIES),]
```
Here I show the top 10 weather events that cause fatalities and injuries:
```{r}
fatal_byType_Sorted[1:10,]
injury_byType_Sorted[1:10,]
```
To get a more intuitive impression, we can make a bar plot showing the fatality and injury numbers caused by different events:
```{r}
par(mfrow = c(1, 2))
par(mar = c(10, 4, 4, 2), cex = 0.8, cex.main = 1.2, cex.lab = 1.2)
barplot(fatal_byType_Sorted$FATALITIES[1:10], names.arg = fatal_byType_Sorted$EVTYPE[1:10], col = 'blue',
        main = 'Top 10 Weather Events for Fatalities', ylab = 'Number of Fatalities')
barplot(injury_byType_Sorted$INJURIES[1:10], names.arg = injury_byType_Sorted$EVTYPE[1:10], col = 'green',
        main = 'Top 10 Weather Events for Injuries', ylab = 'Number of Injuries')
```

It's clear from the plots above, that **TORNADO** is most harmful to population health. The number of fatalities and injuries caused by **TORNADO** is far higher than the other events.

### Q2: Across the United States, which types of events have the greatest economic consequences?
The same way, we first need to extract the relevant data, which are "PROPDMG" (property damages) and "CROPDAMGE" (crop damages). Since these two factors have equivalent importance, in addition to see them seperately, I will also add them up to see a total damage to the economy. Still,  we can sum for those numbers by event types and sort them in a descending order:
```{r}
## Exxtracting and sum over types
propertyDMG_byType <- aggregate(PROPDMG ~ EVTYPE, stormData, sum)
cropDMG_byType <- aggregate(CROPDMG ~ EVTYPE, stormData, sum)
totalDMG_byType <- merge(propertyDMG_byType, cropDMG_byType, by = "EVTYPE")
totalDMG_byType$TOTALDMG <- totalDMG_byType$PROPDMG + totalDMG_byType$CROPDMG
## Sorting
propertyDMG_byType_Sorted <- propertyDMG_byType[order(-propertyDMG_byType$PROPDMG),]
cropDMG_byType_Sorted <- cropDMG_byType[order(-cropDMG_byType$CROPDMG),]
totalDMG_byType_Sorted <- totalDMG_byType[order(-totalDMG_byType$TOTALDMG),]
```
Here I show the top 10 weather events that cause property damages, crop damages and total damages:
```{r}
propertyDMG_byType_Sorted[1:10,]
cropDMG_byType_Sorted[1:10,]
totalDMG_byType_Sorted[1:10,]
```
To get a more intuitive impression, we can make a bar plot showing the damages caused by different events:
```{r}
par(mfrow = c(1, 3))
par(mar = c(10, 4, 4, 2), cex = 0.8, cex.main = 1.2, cex.lab = 1.2)
barplot(propertyDMG_byType_Sorted$PROPDMG[1:10], names.arg = propertyDMG_byType_Sorted$EVTYPE[1:10], col = 'blue',
        main = 'Top 10 Property damages', ylab = 'Property Damages')
barplot(cropDMG_byType_Sorted$CROPDMG[1:10], names.arg = cropDMG_byType_Sorted$EVTYPE[1:10], col = 'green',
        main = 'Top 10 Crop damages', ylab = 'Crop Damages')
barplot(totalDMG_byType_Sorted$TOTALDMG[1:10], names.arg = totalDMG_byType_Sorted$EVTYPE[1:10], col = 'orange',
        main = 'Top 10 Total damages', ylab = 'Total Damages')
```

As shown in the plots, **TORNADO** causes the most property damages, while **HAIL** causing the most crop damages. Overall, **TORNADO** is still the most harmful weather event that causes economic consequences.